### 2022-08-08 description.tex
Work on description, add d_seq to results

% Evolution FST
Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
The Evolution FST is formed by composing a substitution FST that encodes a 64x64
codon model and an indel FST that models insertions and deletions, including
frameshifts.
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model with gaps that can occur at any
position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represents a pairwise alignment.
% The shortest path through the FST corresponds with the best alignment.
However, composing large FSTs is expensive and can be prohibitive \green{(specify
I talk about time)}.
Despite the existence of efficient C++ FST libraries (e.g. openFST
\cite{allauzen2007openfst}), runtime is still limiting for sequence longer than
a few thousand nucleotides.
\green{To solve this issue, the search for an optimal path (alignment)} can be
reformulated as a dynamic programming problem.

>>>>

% Evolution FST
Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
The Evolution FST is formed by composing a substitution FST that encodes a 64x64
codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models
insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model that allows stop codons with gaps that
can occur at any position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represents a pairwise alignment.
% The shortest path through the FST corresponds with the best alignment.
However, composing large FSTs is expensive and can be prohibitive \green{(specify
I talk about time)}.
Despite the existence of efficient C++ FST libraries (e.g. openFST
\cite{allauzen2007openfst}), runtime is still limiting for sequence longer than
a few thousand nucleotides.
\green{To solve this issue, the search for an optimal path (alignment)} is
reformulated as a dynamic programming problem.

### 2022-08-08 description2.tex
Work on description, add d_seq to results

% Evolution FST
Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
This FST is formed by composing a substitution FST that encodes a 64x64 codon
model and an indel FST that models insertions and deletions, including frameshifts.
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model with gaps that can occur at any
position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represent a pairwise alignment.
% To find the most \green{probable/likely} alignment we must find the shortest
% path through.
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
However, composing large FSTs is an expensive operation and can be prohibitive.
Despite the existence of efficient C++ FST libraries (e.g. openFST
\cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
than a few thousand nucleotides each.
To solve this issue, the search for an optimal path (alignment) can be
reformulated as a dynamic programming problem.
% Maintaining (or approximate??) the statistical framework, COATi implements a
% Gotoh-inspired algorithm thus reducing the problem to O(nm) runtime, n and m 
% length of sequences \green{citation?}

% Alignment itself?

% Evolution FST + Q matrix
% Marginal model? - Probably not
% DP - mention in a sentence or two.
% 1 figure: FST model? - results table?

>>>>

% Evolution FST
Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
The Evolution FST is formed by composing a substitution FST that encodes a 64x64
codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models
insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model that allows stop codons with gaps that
can occur at any position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represent a pairwise alignment.
% To find the most \green{probable/likely} alignment we must find the shortest
% path through.
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
All FST operations including creation of models, composition, search for the
shortest path, and other optimization algorithms are performed using the C++
openFST library \parencite{allauzen2007openfst}.

% However, composing large FSTs is an expensive operation and can be prohibitive.
% Despite the existence of efficient C++ FST libraries (e.g. openFST
% \cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
% than a few thousand nucleotides each.
% To solve this issue, the search for an optimal path (alignment) is
% reformulated as a dynamic programming problem.

% Evolution FST + Q matrix
% Marginal model? - Probably not
% DP - probably not
% 1 figure: FST model? - results table?

### 2022-08-08 pairwise_appnote.tex
Work on description, add d_seq to results

% \abstract{Abstract}
\begin{abstract}
\noindent \textbf{Summary:} COATi is a statistical codon-aware pairwise aligner
that supports complex insertion-deletion models and is able to handle artifacts
present in genomic data.\\  % abstract summary can be longer.
\textbf{Availability:} The source code, along with documentation, for COATi is
freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
implemented in C++.\\
\textbf{Supplementary information:} %\green{TODO}.
\end{abstract}

@@

The author(s) would like to thank \green{person1}, \green{person2}, and
\green{number} reviewers for their helpful suggestions.

>>>>

% \abstract{Abstract}
\begin{abstract}
\noindent \textbf{Summary:} COATi is a statistical codon-aware pairwise aligner
that supports complex insertion-deletion models and is able to handle artifacts
present in genomic data.\\  % abstract summary can be longer.
\textbf{Availability:} The source code for COATi, along with documentation, is
freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
implemented in C++.\\
\textbf{Supplementary information:} %\green{TODO}.
\end{abstract}

@@

% The authors would like to thank \green{person1}, \green{person2}, and
% \green{number} reviewers for their helpful suggestions.

### 2022-08-08 results.tex
Work on description, add d_seq to results

COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments,
and had a higher positive (97.3\%) selection accuracy.
MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
average alignment error an order of magnitude larger than COATi and the accuracy
retrieving positive selection of 81.5\% and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
obtained similar results with an average alignment error two orders of magnitude
larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
The accuracy retrieving negative selection was similar across all five
aligners.

>>>>

COATi was significantly more accurate, $d_{seq}$ value of $6$x$10^{-4}$  than
other aligners; all p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments,
and had a higher positive selection accuracy (97.3\%).
MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
average alignment error an order of magnitude larger than COATi ($6$x$10^{-3}$)
and the accuracy retrieving positive selection of 81.5\% and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
obtained similar results with an average alignment error two orders of magnitude
larger than COATi ($0.01$) and a 87.3\% and 69.1\% accuracy retrieving positive
selection.
The accuracy retrieving negative selection was similar across all five
aligners (98.6\% $\pm$ 1.25\%).
